### 2011-11-06 doc/pldi2012/experiments.tex
finished first draft of experiments for first two experiments

To measure Legion's ability to distribute work across a distributed machine, 
three benchmarks were developed and run on a small cluster under a variety
of configurations.  The cluster consists of four systems connected by a 40~Gb/s
Infiniband fabric.  Each system is powered by 2 Intel Xeon X5680 CPUs, each
with 6 cores, and 2 NVIDIA Tesla C2070 GPUs.  The systems were running Linux
with the standard OFED software stack, with GASnet using its ``ibv'' conduit
to run directly on top.

@@

The structure of the circuit simulator was presented in section~\ref{sec:ex}.
Our implementation uses CUDA to perform the computations on the GPU, and Legion
automatically handles the movement of data between the GPUs' memories and the
system memory, but only when necessary.  In fact, with the help of an
application-specific mapper, it is able to keep much of
the data permanently resident in a given GPU for the entirety of the simulation.
Figure~\ref{sfig:results1:perf} shows the relative speedup as the simulation is
spread over multiple nodes in the cluster.  The size of the circuit
being simulated is not changed as the node count increases.  This is therefore
a demonstration of ``strong scaling''.  Adding a second node to the simulation
speeds it up by 77\%, and with four nodes, the simulation is sped up by a total
of 216\% (i.e. a factor of 3.16x).

The bar graph in figure~\ref{sfig:results1:timers} breaks down the simulation's
run time by where it was spent.  The times are summed over all threads, so 
perfect linear scaling would keep the bars flat in this graph.  The total
overhead introduced by Legion is XX \% for the single-node case.
shows a breakdown of the
total amount of time spent

@@

To show that Legion remains efficient for regular partitioning cases as well,
we have implemented a particle simulation benchmark, based heavily on the
{\tt fluidanimate} benchmark from the PARSEC suite.  The simulation uses a
cyclic 2D grid of cells instead of PARSEC's 3D grid with boundaries, and 
The fluid benchmark used here differs slightly from the PARSEC version.  It
uses a 2D grid of cells instead of a 3D grid, but the computations being 
performed are preserved.
Although the 
simulation is moving and colliding individual particles, those particles
are placed in cells that make up a regular grid, and the time steps and grid
sizes are chosen carefully to ensure that a particle can only interact with
particles in the same cell or in one immediately adjacent.  The grid is
divided into blocks to be processed by individual threads, with only the
particles in the edge cells being of interest to other threads.  The PARSEC
implementation uses fine-grained locking and the assumption of cache
coherence to access neighboring particles, but neither of these are available
across a cluster.  Our fluid benchmark replaces these accesses with rings of
explicit edge, or \emph{ghost}, cells that are exchanged between adjacent
blocks several times in each iteration.  The capturing of these inter-block
interactions through edge cells and the partitioning of those cells into
subregions shoulds the runtime the ability to schedule data movement in
advance and do a more precise enforcement of ordering constraints (e.g.
requiring a task to wait only for the tasks in the previous pass that computed
one of the ghose cells this task needs).

Figure~\ref{sfig:results2:perf} shows the performance of the fluid benchmark
relative to the serial Legion version.  MORE TEXT HERE.

The bar chart in figure~\ref{sfig:results2:timers} shows MORE STUFF.

>>>>

To measure Legion's ability to distribute work across a distributed machine, 
three benchmarks were developed and run on a small cluster under a variety
of configurations.  The cluster consists of four systems connected by a 40~Gb/s
Infiniband fabric.  Each system is powered by 2 Intel Xeon X5680 CPUs, each
with 6 cores, and 2 NVIDIA Tesla C2070 GPUs.  The systems were running Linux
with GASNet using the {\tt ibv} conduit to interface directly with the standard
OFED software stack for low-latency access to the Infiniband hardware.

@@

The structure of the circuit simulator was presented in Section~\ref{sec:ex}.
Our implementation uses CUDA to perform the computations on the GPU, and Legion
automatically handles the movement of data between the GPUs' memories and the
system memory, but only when necessary.  In fact, with the help of an
application-specific mapper, it is able to keep much of
the data permanently resident in a given GPU for the entirety of the simulation.
Figure~\ref{sfig:results1:perf} shows the relative speedup as the simulation is
spread over multiple nodes in the cluster.  The size of the circuit
being simulated is not changed as the node count increases.  This is therefore
a demonstration of ``strong scaling''.  Adding a second node to the simulation
speeds it up by 86\%, and with four nodes, the simulation is sped up by a total
of 241\% (i.e. a factor of 3.41x).  When enabling the use of both GPUs in each
node (a decision that requires no changes to the application), the we achieve
a speedup of 5.90x when using all eight GPUs.

The dip in the performance graph for the 6 GPU case is an artifact of the
application code's decision to partition the circuit into 8 pieces.  The
current runtime interface does not allow the application to make queries
regarding the machine configuration, so it is not able to adapt to
configurations that differ from the initial design target.  The Legion runtime
will always guarantee correct execution, but performance can suffer if the
tasks provided by the application do not divide evenly over the machine.

Table\~ref{sfig:results1:timers} breaks down the execution time by where it was
spent.  The times are summed over all threads, so the aggregate execution time
will always exceed the elapsed wall clock time.  The time spent in the 
application code is nearly constant in all configurations, but the time spent
copying data jumps as soon as the simulation is spread across multiple nodes.
This is not unexpected, and while the the overhead is responsible for the
less-than-linear performance scaling, the slow growth in the absolute overhead
(after the initial jump) suggests that larger clusters would enjoy additional
speedup, even without increasing the size of the problem being tackled.

@@

To show that Legion remains efficient for regular partitioning cases as well,
we have implemented a particle simulation benchmark based heavily on the
{\tt fluidanimate} benchmark from the PARSEC \cite{bienia11benchmarking} suite.
Our simulation uses a cyclic 2D grid of cells instead of PARSEC's 3D grid with
boundaries, but the computations being performed are preserved.
Although the 
simulation is moving and colliding individual particles, those particles
are placed in cells that make up a regular grid.  The grid is
divided into blocks to be processed by individual threads, with only the
particles in the edge cells being of interest to other threads.  The PARSEC
implementation uses fine-grained locking and the assumption of cache
coherence to access neighboring particles, but neither of these are conducive
to performance scaling across a cluster.  Our fluid benchmark replaces these
accesses with rings of
explicit edge, or \emph{ghost}, cells that are exchanged between adjacent
blocks several times per iteration.  The capturing of these inter-block
interactions through edge cells and the partitioning of those cells into
subregions provides the runtime the ability to schedule data movement in
advance and perform a more precise enforcement of the ordering
constraints necessary for correctness (e.g.
requiring a task to wait only for the subset of tasks in the previous pass 
that computed one of the ghost cells being read by this task).

Figure~\ref{sfig:results2:perf} shows the throughput of the fluid benchmark when
run on various subsets of the CPUs and nodes in the cluster.  Performance is
normalized with respect to the serial execution case.  As in the circuit
simulation case, the same size problem is used for all runs, and we see 
speedups of up to 10.14x when using the entire cluster (a total of 48 CPUs).
The execution time breakdown in Table~\ref{sfig:results2:timers} shows some
differences in scaling behavior compared to the circuit simulation.  The time
spent performing copies is much smaller on absolute scale, but shows the same
large step when leaving a single node, but then relative insensitivity to the
number of nodes.  This indicates that the runtime is able to send data only to
the task that needs it rather than having to perform conservative broadcasts of
data to maintain consistency.

A more qualitative difference between the two simulations is that fluid shows
significant variability in the time spent in the application code, and there
are two separate issues at work.  First, the runtime is taking advantage of
the CPU's ability to perform RDMA operations to place some of the shared data
in the remotely-accessible memory windows provided by GASNet.  Although access
to this memory is significantly slower than accesses to a node's system memory,
being able to avoid copies entirely in some cases is a net win.

The second significant jump in application time is when using all 12 CPUs in
each node.  With application tasks running on all the available CPU cores,
there are no ``extra'' cores for running the OS, device drivers, and the like.
For example, GASNet's low-latency implementation counts on a CPU core being
available for 
a thread to spin on message arrival notifications.  Allowing that thread to
sleep (and then wait for the OS rescheduling delay) degrades system performance
much more than the penalty being observed in our fluid simulation.  Now that
one can generally count on having many CPU cores in any new system, it might
make sense to reserve one (or more) in the low-level runtime for background
tasks like these in order to provide more predictable performance on the CPU
cores that are exposed to the high-level runtime.
